Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices

### 2015-09-22 flux-entropy-paper/flux-entropy-paper.tex
Added apostrophe
I added an apostrophe to show that I can use github! It took me ages to
find the commit button! I am still having troubles running latex; the
github directory is not self contained; you guys have different latex
configurations to me.
finished more Ben edits

\begin{abstract}
  In this note we describe the analysis of a data archive for an
  ensemble musical performance practice on touch screens where
  transition matrix measures are able to distinguish between sessions
  of different contexts, and the internal structure of these sessions.
  Traditional methodologies for studying interfaces that support
  collaborative creativity can be time-consuming and intrusive to
  implement but our methods may provide a practical alternative for
  large scale analysis of creative interactions.

  Our archive includes 95 touch data logs of ensemble touch-screen
  interactions in concerts, rehearsals and formal HCI studies of
  groups of musicians over more than two years. The data analysis has
  been undertaken by first classifying these data into higher level
  gesture types and summarising the performers' transitions between
  these gestures. We find that two transition-matrix measure, flux,
  and entropy, can be used to distinguish different performance types
  and contexts, as well as the internal structure (beginning, middle
  and end) of the performances in our corpus.
\end{abstract}

\keywords{  
  Creativity Support Tools;
  Agent; 
  Design;
  Methodology}

\category{H.5.5.}{Information Interfaces and Presentation
  (e.g. HCI)}{Sound and Music Computing}{ Systems}

@@

Modern computing devices and interfaces continue to present
opportunities to support new kinds of creative expression and
collaborative interaction~\cite{Resnick:2005yu}. However, it has been
acknowledged that understanding and evaluating interfaces in creative
contexts can be challenging~\cite{Shneiderman:2007qv}. In the context
of live performance, interface evaluation using questionnaires, or
ethnographic data gathering can miss the dynamic needs of artists
during collaborative interactions and may be impractical in live
performative contexts. When performers use a combuter-based artistic
interface, logged interaction data which has been collected for
curation and preservation~\cite{England:2014ys}, might also allow
quantitative investigation into these performances. In this note we
will show how an archive of collaborative performance data logs can be
subjected to a quantitative analysis based on transition probabilities
between gestural states. We will demonstrate that two measures on
transition matrices, flux and entropy, show significant differences
between different kinds of performances, and give insight into the
internal structure of these performances.

Our archive of interaction data logs has been accumulated to support a
long-term practice-led research program into simple iPad interfaces
for group music-making similar to those described in Martin et
al~\cite{Martin:2014cr}. Over a period of more than two years (April
2013--June 2015), we collected data from group music-making sessions
where the participants used our custom-built, touch-screen apps on
Apple iPad devices. Although several different apps have been included
in our corpus, they shared a common mode of free-form touch
interaction where tapping produced short sounds and swirling or
swiping produced long sounds. Apart from this commonality, the apps
featured a variety of sound palettes, different kinds of visual
feedback, and different kinds of networked interactions (between the
iPads in an ensemble and with a central server).

@@

Representative photographs of some of these session contexts are are
shown Figure~\ref{fig:three-performance-contexts}) and other
descriptive statistics about the sessions are shown in Table
\ref{corpus-table}. In our corpus, the median length of sessions is
7m26s and the median number of participants per ensemble is four. Each
session record is labelled with metadata about the session
\emph{context} (performance, rehearsal, study), session \emph{type}
(improvisation or composition), and \emph{instrumentation}
(iPads-only, or iPads with acoustic percussion). The distribution of
recorded sessions by context and type is shown in Figure
\ref{fig:count-data}.

@@

In our corpus of iPad musical performances, logs of low-level touch
data record each performers micro-gestures. However, it has been
established that musical performances on touch-screens can be seen in
terms of vocabularies of continuous macro gestures that unfold over
several seconds of touch interaction~\cite{Martin:2014cr}. Improvised
performances can be then thought of as a sequence of such
states~\cite{Pressing:1988uo} for each performer.

@@

In this note we have extended previous work using transition matrices
to understand HCI processes with the application of two simple matrix
measures, flux and entropy, both of which have been demonstrated to
have significant discriminatory power with regards to a corpus of 95
collaborative touch-screen interactions. Not only do these measures
differentiate between different types of sessions and the internal
structure of these sessions, but they have explanatory power that has
thrown new light on the musical interactions under examination. It is
notable that, for improvised performances, live performances had the
lowest measure in both flux and entropy, while lab-studies had the
highest. This may be due to the increased pressure on performers when
improvising in front of an audience. The internal investigation of
performances session revealed more structural variety in
improvisations than compositions. In response, new works could be
composed that specifically explore different gestural structures in
touch-screen performance

HCI researchers are increasingly concerned with interactions that
happen in-the-wild, where the creative collaborations of many users
may be recorded, but where conducting surveys or interviews would be
innapropriate. We suggest that analysing interaction-state transitions
using matrices and simple measures can lead to new insights into HCI
interactions outside the lab.

Data-mining logs of interactions with creative interfaces sugggest
further approaches for future work. The macro-gestures that these
processes reveal could be used as the basis for new works of new media
art~\cite{Manovich:2002ly}. Additionally, a wider dissemination of
such log data could better support the replication of findings in HCI
research~\cite{Wilson:2011ve}

>>>>

\begin{abstract}

  In this note we describe the analysis of an interaction data archive
  associated with a long-term practice-led research program
  investigating ensemble musical performance practice on touch
  screens. We use transition matrix measures to distinguish between
  sessions of different contexts, and the internal structure of these
  sessions. The data analysis proceeds by first classifying these data
  into higher level gesture types and summarising the performers'
  transitions between these gestures. We find that two
  transition-matrix measure, flux, and entropy, can be used to
  distinguish different performance types and contexts, as well as the
  internal structure of the performances in our corpus.
\end{abstract}

\keywords{  
  Creativity Support Tools;
  Collaborative Interaction;
  Methodology}

\category{H.5.5.}{Information Interfaces and Presentation
  (e.g. HCI)}{Sound and Music Computing}{Systems}

@@

Modern computing devices and interfaces continue to present
opportunities to support new kinds of creative expression and
collaborative interaction~\cite{Resnick:2005yu}. However, it has been
acknowledged that understanding and evaluating interfaces in creative
contexts can be challenging~\cite{Shneiderman:2007qv}. This problem is
even more pronounced in artistic \emph{performance}, where subjective
interface evaluation through questionnaires or interviews may be both
practically difficult and culturally inappropriate. When the interface
is digital, however, it is still usually trivial to log interaction in
these situations, which is useful for curation and
preservation~\cite{England:2014ys}. The question is: can this log data
be used not just as a record for posterity, but in support of a
broader creative-practice led research program?

In this note we will show how an archive of collaborative performance
data logs can be subjected to a quantitative analysis based on
transition probabilities between gestural states. We will demonstrate
that two measures on transition matrices, flux and entropy, show
significant differences between different kinds of performances, and
give insight into the internal structure of these performances.

Our archive of interaction data logs has been accumulated to support a
long-term practice-led research program into simple iPad interfaces
for group music-making similar to those described in Martin et
al~\cite{Martin:2014cr}. Over a period of more than two years (April
2013--June 2015), we collected data from group music-making sessions
where the participants used our custom-built, touch-screen apps on
Apple iPad devices. Although several different apps have been included
in our corpus, they shared a common mode of free-form touch
interaction where tapping produced short sounds and swirling or
swiping produced long sounds. Apart from these similarities, the apps
featured a variety of sound palettes, different kinds of visual
feedback, and different kinds of networked interactions (between the
iPads in an ensemble and with a central server).

@@

Some of these session contexts are are shown
Figure~\ref{fig:three-performance-contexts}) and other descriptive
statistics about the sessions are shown in Table \ref{corpus-table}.
In our corpus, the median length of sessions is 7m26s and the median
number of participants per ensemble is four. Each session record is
labelled with metadata about the session \emph{context} (performance,
rehearsal, study), session \emph{type} (improvisation or composition),
and \emph{instrumentation} (iPads-only, or iPads with acoustic
percussion). The distribution of recorded sessions by context and type
is shown in Figure \ref{fig:count-data}.

@@

In our corpus of iPad musical performances, logs of low-level touch
data record each performer's micro-gestures. However, it has been
established that musical performances on touch-screens can be seen in
terms of vocabularies of continuous macro gestures that unfold over
several seconds of touch interaction~\cite{Martin:2014cr}. Improvised
performances can be then thought of as a sequence of such
states~\cite{Pressing:1988uo} for each performer.

@@

In this note we have extended previous work using transition matrices
to understand HCI processes with the application of two simple matrix
measures, flux and entropy, both of which have been demonstrated to
have significant discriminatory power with regards to a corpus of 95
collaborative touch-screen interactions. Not only do these measures
differentiate between different types of sessions and the internal
structure of these sessions, they provide qualitative insight into the
performances as well. For improvised performances, live performances
had the lowest measure in both flux and entropy, while lab-studies had
the highest. This may be due to the increased pressure on performers
when improvising in front of an audience. The internal investigation
of performances session revealed more structural variety in
improvisations than compositions. In response, new works could be
composed that specifically explore different gestural structures in
touch-screen performance.

Data-mining logs of interactions with creative interfaces suggest
further approaches for future work. The macro-gestures that these
processes reveal could be used as the basis for new works of new media
art~\cite{Manovich:2002ly}. Additionally, a wider dissemination of
such log data could better support the replication of findings in HCI
research~\cite{Wilson:2011ve}. Finally, flux and entropy proved useful
in the analysis of our data corpus, but other transition matrix
metrics may be more suitable for different interaction data.

HCI researchers are increasingly concerned with interactions that
happen in-the-wild, where the creative collaborations of many users
may be recorded, but where conducting surveys or interviews is
innapropriate. We suggest that analysing interaction-state transitions
using matrices and simple measures can lead to new insights into HCI
interactions outside the lab.
